---
title: "Helsinki open data R tools"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Helsinki open data R tools}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE, 
  warning = FALSE,
  fig.height = 7, 
  fig.width = 7,
  dpi = 75
)
```


helsinki - tutorial
===========

This R package provides tools to access open data from the Helsinki region in Finland
as part of the [rOpenGov](http://ropengov.org) project.

For contact information and source code, see the [github page](https://github.com/rOpenGov/helsinki). 

## Available data sources

[Helsinki region district maps](#aluejakokartat) (Helsingin seudun aluejakokartat)

* Aluejakokartat: kunta, pien-, suur-, tilastoalueet (Helsinki region district maps)
* Äänestysaluejako: (Helsinki region election district maps)
* Source: [Helsingin kaupungin Kiinteistövirasto (HKK)](http://ptp.hel.fi/avoindata/)

Helsinki Real Estate Department (HKK:n avointa dataa)

* Spatial data from [Helsingin kaupungin Kiinteistövirasto (HKK)](http://ptp.hel.fi/avoindata/) availabe in the [gisfin](https://github.com/rOpenGov/gisfin) package, see [gisfin tutorial](https://github.com/rOpenGov/gisfin/blob/master/vignettes/gisfin_tutorial.md) for examples

[Helsinki region environmental services](#hsy) (HSY:n avointa dataa)

* Väestötietoruudukko (population grid)
* Rakennustietoruudukko (building information grid)
* SeutuRAMAVA (building land resource information(?))
* Source: Helsingin seudun ympäristöpalvelut, HSY

[Service and event information](#servicemap)

* [Helsinki region Service Map API](http://api.hel.fi/servicemap/v1/) (Pääkaupunkiseudun Palvelukartta)
* [Helsinki Linked Event API](http://api.hel.fi/linkedevents/v0.1/)

[Helsinki Region Infoshare statistics API](#hri_stats)

* [Aluesarjat (original source)](http://www.aluesarjat.fi/) (regional time series data)
* Source: [Helsinki Region Infoshare statistics API](http://dev.hel.fi/stats/)


List of potential data sources to be added to the package can be found [here](todo-datasets.Rmd).

## Installation

Release version for general users:

```{r install_default, eval=FALSE}
install.packages("helsinki")
```

Development version for developers:

```{r install_remotes, eval=FALSE}
library(remotes)
remotes::install_github("ropengov/helsinki")
```

Load package.

```{r load, message=FALSE, warning=FALSE, results='hide'}
library(helsinki)
```

## <a name="aluejakokartat"></a>Helsinki region district maps

Helsinki region district maps (Helsingin seudun aluejakokartat) and election maps (äänestysalueet) from [Helsingin kaupungin Kiinteistövirasto (HKK)](http://ptp.hel.fi/avoindata/) are available in the helsinki package with `data(aluejakokartat)`. The data are available as both spatial object (`sp`) and data frame (`df`). These are preprocessed in the [gisfin](https://github.com/rOpenGov/gisfin) package, and more examples can be found in the [gisfin tutorial](https://github.com/rOpenGov/gisfin/blob/master/vignettes/gisfin_tutorial.md). 

```{r aluejako_data, message=FALSE}
# Load aluejakokartat and study contents
data(aluejakokartat)
str(aluejakokartat, m=2)
```


## <a name="servicemap"></a>Service and event information

Function `get_servicemap()` retrieves regional service data from the new [Service Map API](http://api.hel.fi/servicemap/v2/), that contains data from the [Service Map](http://dev.hel.fi/servicemap/).

```{r servicemap, message=FALSE, warning=FALSE}
# Search for "puisto" (park) (specify q="query")
search_puisto <- get_servicemap(query="search", q="puisto")
# Study results: 47 variables in the data frame
str(search_puisto, max.level = 1)
```

We can see that this search returns a large number of results, over 2000. The results are returned as pages, where each page has 20 results by default. By giving no additional search parameters, we get 20 results from the first page of search results.

```{r}
# Get names for the first 20 results
search_puisto$results$name$fi

# See what kind of data is given for services
names(search_puisto$results)
```

More results could be retrieved and viewed by giving additional search parameters.

```{r}
search_puisto <- get_servicemap(query="search", q="puisto", page_size = 30, page = 2)
search_puisto$results$name$fi
```

As we could see from above example, the returned data frame had 47 observations. At full width this output can be messy to handle. One option would be to turn it into a tibble, the other to limit the extent of the query at the start. This can be done with only parameter:

```{r}
# Search for padel-related services in Helsinki
search_padel <- get_servicemap(query="search", input="padel", only="unit.name, unit.location.coordinates, unit.street_address", municipality="helsinki")
search_padel$results
```

Function `get_linkedevents()` retrieves regional event data from the new [Linked Events API](http://api.hel.fi/linkedevents/v0.1/).

```{r linkedevents, message=FALSE, warning=FALSE}
# Search for current events
events <- get_linkedevents(query="event")
# Get names for the first 20 results
events$data$name$fi
# See what kind of data is given for events
names(events$data)
```



## <a name="hri_stats"></a> Helsinki Region Infoshare statistics API

Function `get_hri_stats()` retrieves data from the [Helsinki Region Infoshare statistics API](http://dev.hel.fi/stats/).

```{r hri_stats1, message=FALSE, warning=FALSE}
# Retrieve list of available data
stats.list <- get_hri_stats(query="")
# Show first results
head(stats.list)
```

Specify a dataset to retrieve. The output is currently a three-dimensional array.

```{r hri_stats2, message=FALSE, warning=FALSE}
# Retrieve a specific dataset
stats.res <- get_hri_stats(query=stats.list[1])
# Show the structure of the results
str(stats.res)
```

The implementation will be updated and more examples will be added in the near future.


### Citation

**Citing the data:** See `help()` to get citation information for each data source individually.

**Citing the R package:**

```{r citation, comment=NA}
citation("helsinki")
```

### Session info


This vignette was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```

