---
title: "Helsinki open data R tools"
author: "Pyry Kantanen"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Helsinki open data R tools}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE, 
  warning = FALSE,
  fig.height = 7, 
  fig.width = 7,
  dpi = 75
)
```


helsinki - tutorial
===========

This R package provides tools to access open data from the Helsinki region in Finland
as part of the [rOpenGov](http://ropengov.org) project.

For contact information and source code, see the project's [GitHub page](https://github.com/rOpenGov/helsinki).

## Installation

Release version for most users:

```{r install_default, eval=FALSE}
install.packages("helsinki")
```

Development version for developers and other interested parties:

```{r install_remotes, eval=FALSE}
library(remotes)
remotes::install_github("ropengov/helsinki")
```

Load the package.

```{r load, message=FALSE, warning=FALSE, results='hide'}
library(helsinki)
```

## API Access

The package has basic functionalities for interacting with WFS APIs. 

All available features of a given API can be easily listed with `get_feature_list()`:

```{r wfs1, eval = TRUE}
url <- "https://kartta.hsy.fi/geoserver/wfs"

hsy_features <- get_feature_list(base.url = url)
feature_of_interest <- hsy_features$Name[20]
# Location of waterposts
feature_of_interest
```

When the wanted feature and its Name (in other words: Namespace:Title combination) is known, it can be downloaded with `get_feature()` by providing the correct base.url and the Name as the typename parameter.

```{r wfs2, eval = TRUE}
# downloading a feature
waterposts <- get_feature(base.url = url, typename = feature_of_interest)
# Visualizing the location of waterposts
plot(waterposts$geom)
```

Dots on a blank canvas do not make much sense and therefore helsinki-package has `get_city_map()` function for downloading city district boundaries. An example of this is provided in the [Helsinki region district maps](#helsinki-region-district-maps) section of this vignette.

Helsinki-package provides an easy-to-use menu-driven `select_feature()` function that effectively combines `get_feature_list()` and `get_feature()`. At default it only returns the Name of the wanted function, but if download parameter is set to TRUE, it returns an sf_object which can be easily visualized.

```{r wfs3, eval = FALSE}
# Standard use
selected_feature <- select_feature(base.url = url)
feature <- get_feature(base.url = url, typename = selected_feature)

# Skipping a redundant step with parameter get = TRUE
feature <- select_feature(base.url = url, get = TRUE)
```

### Helsinki Region Environmental Services HSY open data

The above example shows a general use case which can easily be applied to Helsinki Region Environmental Services (HSY) WFS API as well as other service providers' APIs. 

For legacy reasons, helsinki-package has also some specialized functions that aim to make downloading often used data as easy as possible. 

Specifically, there are two new functions that replace deprecated functionalities from `get_hsy()` function: `get_vaestotietoruudukko()` (population grid) and `get_rakennustietoruudukko()` (building information grid). As of writing, years 2015 to 2019 are supported but the API may be updated at any time and `get_feature_list()` can be used to download datasets that are not baked into these functions.

```{r get_hsy_examples, eval = TRUE}
pop_grid <- get_vaestotietoruudukko(year = 2018)
building_grid <- get_rakennustietoruudukko(year = 2016)

library(ggplot2)

# Logarithmic scales to make the visualizations more discernible
ggplot(pop_grid) + geom_sf(aes(colour=log(asukkaita), fill=log(asukkaita)))
ggplot(building_grid) + geom_sf(aes(colour=log(kerala_yht), fill=log(kerala_yht)))
```

While easy enough to build, specialized functions such as these are probably not something that power users want to rely on in their work flows. They also add more manual phases to package maintenance and therefore are probably not the direction we're heading in the future. If you feel differently about this and there is a dataset that gets a lot of use, feel free to drop us a suggestion in [GitHub](https://github.com/rOpenGov/helsinki/issues).

## Service and event information

Function `get_servicemap()` retrieves regional service data from city of Helsinki [Service Map API](http://api.hel.fi/servicemap/v2/), that contains data from the [Service Map](http://dev.hel.fi/servicemap/).

```{r servicemap, message=FALSE, warning=FALSE}
# Search for "puisto" (park) (specify q="query")
search_puisto <- get_servicemap(query="search", q="puisto")
# Study results: 47 variables in the data frame
str(search_puisto, max.level = 1)
```

We can see that this search returns a large number of results, over 2000. The results are returned as pages, where each page has 20 results by default. By giving no additional search parameters, we get 20 results from the first page of search results.

```{r}
# Get names for the first 20 results
search_puisto$results$name$fi

# See what kind of data is given for services
names(search_puisto$results)
```

More results could be retrieved and viewed by giving additional search parameters.

```{r}
search_puisto <- get_servicemap(query="search", q="puisto", page_size = 30, page = 2)
search_puisto$results$name$fi
```

As we could see from above example, the returned data frame had 47 observations. At full width this output can be messy to handle. One option would be to turn it into a tibble, the other to limit the extent of the query at the start. This can be done with only parameter:

```{r}
# Search for padel-related services in Helsinki
search_padel <- get_servicemap(query="search", input="padel", only="unit.name, unit.location.coordinates, unit.street_address", municipality="helsinki")
search_padel$results
```

Function `get_linkedevents()` retrieves regional event data from the new [Linked Events API](http://api.hel.fi/linkedevents/v0.1/).

```{r linkedevents, message=FALSE, warning=FALSE}
# Search for current events
events <- get_linkedevents(query="event")
# Get names for the first 20 results
events$data$name$fi
# See what kind of data is given for events
names(events$data)
```

## Helsinki region district maps

Helsinki region geographic data can be accessed from a WFS API by using the get_city_map() function. Data is available for all 4 cities in the capital region: Helsinki, Espoo, Vantaa and Kauniainen. 

Administrative divisions can be accessed on 3 distinct levels: "suuralue", "tilastoalue" and "pienalue". Literal, completely unofficial translations for these could be "grand district", "statistical area" and "(minor) district". The naming convention of these levels is sometimes confusing even in Finnish documents and different names can vary by city and time. 

The main takeaway is that "suuralue" is the highest-level division and "pienalue" is the most granular level of division. "Tilastoalue" is somewhere between these two.

As promised earlier in [API Access](#api-access), the following example gives an idea on how to visualize waterpost locations (and, you guessed right, other kinds of spatial data) on a map.

```{r maps1, eval = TRUE}
helsinki <- get_city_map(city = "helsinki", level = "suuralue")
espoo <- get_city_map(city = "espoo", level = "suuralue")
vantaa <- get_city_map(city = "vantaa", level = "suuralue")
kauniainen <- get_city_map(city = "kauniainen", level = "suuralue")

library(ggplot2)

ggplot() +
  geom_sf(data = helsinki) +
  geom_sf(data = espoo) +
  geom_sf(data = vantaa) +
  geom_sf(data = kauniainen) +
  geom_sf(data = waterposts)
```

In addition, it is possible to download "aanestysalue" (voting district) divisions for the city of Helsinki. Currently this data is not available for other cities and it must be accessed from other sources.

```{r maps2, eval=FALSE}
library(sf)
map <- get_city_map(city = "helsinki", level = "suuralue")
plot(sf::st_geometry(map))

voting_district <- get_city_map(city = "helsinki", level = "aanestysalue")
plot(sf::st_geometry(voting_district))
```

For other cities than Helsinki voting areas (and older district maps that shouldn't be too different from current maps) can be viewed from the legacy dataset with `data(aluejakokartat)`. The data included with the packaged is available as both spatial object (`sp`) and data frame (`df`). This data was preprocessed with the now-defunct gisfin-package. The current, actively maintained version of gisfin is called [geofi](https://github.com/rOpenGov/geofi)).

The included aluejakokartat-dataset can also be used if there is for some reason a temporary internet outage in your area and you just need to draw maps on Finnish capital region. It will be bundled with the package for the time being.

Here are the contents of `data(aluejakokartat)`:

```{r maps3, message=FALSE}
# Load aluejakokartat and study contents
data(aluejakokartat)
str(aluejakokartat, m=2)
```

## Helsinki Region Infoshare statistics API

Function `get_hri_stats()` retrieves data from the [Helsinki Region Infoshare statistics API](http://dev.hel.fi/stats/).

```{r hri_stats1, message=FALSE, warning=FALSE}
# Retrieve list of available data
stats.list <- get_hri_stats(query="")
# Show first results
head(stats.list)
```

Specify a dataset to retrieve. The output is currently a three-dimensional array.

```{r hri_stats2, message=FALSE, warning=FALSE}
# Retrieve a specific dataset
stats.res <- get_hri_stats(query=stats.list[1])
# Show the structure of the results
str(stats.res)
```

## Licensing and Citations

### Citing the data 

See `help()` to get citation information for each function and related data sources. 

If no such information is explicitly stated, see data provider's website for more information.

### Citing the R package

```{r citation, comment=NA}
citation("helsinki")
```

### Session info

This vignette was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```

