<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{An R Markdown Vignette made with knitr}
-->

Helsinki  R tools
===========

This is an [rOpenGov](http://ropengov.github.com/helsinki) R package
providing tools for open Helsinki data.


### Installation

Release version for general users:

```{r install, eval=FALSE}
install.packages("helsinki")
library(helsinki)
```

Development version for developers:

```{r install2, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("helsinki", "ropengov")
library(helsinki)
```

Further installation and development instructions at the [home
page](http://ropengov.github.com/helsinki).


### Examples

For further usage examples, see
[Louhos-blog](http://louhos.wordpress.com) and
[Datawiki](https://github.com/louhos/sorvi/wiki/Data), and
[takomo](https://github.com/louhos/takomo/tree/master/Helsinki).


### Licensing and Citations

This work can be freely used, modified and distributed under the
[Two-clause FreeBSD
license](http://en.wikipedia.org/wiki/BSD\_licenses). Cite Helsinki R
package and and the appropriate data provider, including a url
link. Kindly cite the R package as 'Leo Lahti, Juuso Parkkinen ja
Joona Lehtomäki (2013). helsinki R package. URL:
http://ropengov.github.io/helsinki)'.


### Session info


This vignette was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```

